Predicting complex diseases: performance and robustness [OF NEW METHOD?]

### 2018-03-12 paper/draft.tex
small English changes

We present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in a previous study. We quickly discard the T-Trees method and show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method are very sensitive to the threshold of inclusion of SNPs, depending on the architecture of disease, as shown in previous studies.\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:}
}

@@

Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs have proven to be useful for different applications, such as finding a common genetic contribution between two diseases \cite[]{purcell2009common}. 
Personalized medicine is another major application of PRSs. Personalized medicine ..will.. use PRSs in screening campaigns in order to identify high-risk individuals for a given disease. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. Moreover, screening based on sequencing seems to make individuals positively change their health behavior while not causing patient anxiety nor depression \cite[]{vassy2017impact}.
Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. 
For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.

Many methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- also sometimes called genetic profiling ..or even just PRS.. -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique only use summary statistics which makes it very suitable and also very fast.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. Yet, these models are not designed for predicting a binary trait such as a disease status and have proven to not be optimal at such task in another comparative study \cite[]{abraham2013performance}. Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRSs for complex human disease by jointly estimating SNP effects. Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.

@@

There have already been comparative studies about predicting human binary phenotypes based on genotype data. 
\citep{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
\cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models performed better in a range of sparse to polygenic disease architectures. 
\cite{spiliopoulou2015genomic} investigated the effect of relatedness in prediction performance.
\cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but ..richer.. than the one used in \cite{zhou2013polygenic}. In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model for generating liability ..scores.., one with only linear effects, and one which combines linear, dominant and interaction effects. First, we quickly discard T-Trees as being the best method for predicting disease status based on genotypes.
Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method are very sensitive to the threshold of inclusion of SNPs, depending on the architecture of disease, as shown in \cite{ware2017heterogeneity}.

@@

We use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. Composition of this dataset is presented in table \ref{tab:celiac-data}.
Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
For simulations, in order to remove the structure induced by the celiac disease status, we ..first keep only.. controls from this cohort. Secondly, in order to remove population structure, we ..keep only.. people from the UK. Thirdly, we further subset these British people based on deviation from the robust Mahalanobis distance on principal components. This 3-step filtering results in 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing''). We decided to get rid of population structure because it can affect the predictive performance of methods, which we will investigate in another study.

@@

To simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% [CITATION?]. We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (..from 1000 to 6000 by 1000..), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part.. of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combines linear, dominant and interaction effects).
For the ``simple model'', we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights (generated with a Gaussian or a Laplace distribution), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs). 
For the ``fancy model'', we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$). We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$. Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still ..display.. a marginal effect. 
For both models, the $y_i$ are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.

We make 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
Scenario \textnumero1 uses the whole dataset (the 22 chromosomes) and a training set of size 6000. It compares all methods (see section \ref{sec:methods}) on all other combination of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}). Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more than one hour of computation for only one simulation. 
Scenario \textnumero2 consists in 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of only the chromosome 6. Reducing the number of SNPs aims at increasing the ..polygenicity.. of our simulated models and at ..virtually.. increasing the sample size [REF? DUDBRIDGE?]. For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
Finally, scenario \textnumero3 reuses the whole dataset but varying the size of the training set in order to assess how the sample size affects predictive performance of the PRS and the ``logit-simple'' methods. 100 simulations are run using 300 causal SNPs anywhere on the genome.

@@

In this study, we use two different measures of predictive accuracy. 
First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,hanley1982meaning,lusted1971signal}. In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
This measure indicates how well we can distinguish between cases and controls using PRSs.
As a second measure, the partial AUC for specificities between 90\% and 100\% is also reported \cite[]{dodd2003partial,mcclish1989analyzing}. This measure is similar to the AUC, but focuses on high specificities (i.e.\ discarding most of the healthy population as being diseased), which is often the most useful part of the ROC curve in clinical settings.

@@

\subsection{Compared methods} \label{sec:methods}

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Studies (GWAS), named summary statistics.
For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with each other. 
Then, thresholding (T) consists in removing SNPs that are within a certain level of significance (p-value threshold to be determined). 
Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T, E_\text{clumping}) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.

For the C+T method, we report three different scores of prediction: one including all the SNPs (remaining after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}). Note that, normally, this optimal threshold would have to be learned on the training set only in order to prevent from over-fitting the data. We consider this reported maximum AUC as an upper bound of the AUC for the C+T method.
We call these three reported scores respectively ``PRS-all'', ``PRS-stringent'' and ``PRS-max'' in the results.

T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}. Since this method is significantly different from the other presented methods and achieved surprisingly high predictive performance in the past, we decided to include T-Trees in this comparison in order to assess whether or not it could give superior predictive performance. We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}. Yet, we use only 100 trees (instead of 1000) because it showed only a very subtle increase of AUC while ..taking way too much time.. (e.g.\ AUC of 81.5\% instead of 81\%, data not shown). %We call this method ``T-Trees'' in the results.

The last method we compare is a penalized logistic regression.
We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined. 
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be exactly zero and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$); 
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we always use the elastic-net regularization with $\alpha = 0.5$, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}. 
This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values). To get an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is chosen. Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ by using the geometric median) or resulting scores can be combined (e.g.\ by averaging the $K$ probabilities for each individual). 
We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making the whole procedure much faster.
Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We call this method ``logit-simple'' in the results.

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct another dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times more variables than the initial one, on which we can apply the penalized logistic regression with the CMSA procedure, as described previously. We call this method ``logit-triple'' in the results.

@@

In ..the.. results, we use estimates of maximum achievable AUC values for the simulations: 84\% (94\%) for heritabilities of 50\% (80\%) based on three estimations (see Supplementary Materials). Note that we do not report the simulation results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).

@@

In scenario \textnumero1, we compared our penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') on the whole genome (Table \ref{tab:simus}). 
Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the C+T method in any scenario (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}). 
The penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}). 
In the case where there are correlation between predictors (i.e.\ when we simulate 30 SNPs in the HLA region), the ``logit-simple'' method particularly performs better than the ``PRS-max'' one, with a mean AUC of 92.5\% versus 83\%, for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
We also report that the choice of the threshold of inclusion of SNPs in the C+T method can greatly impact its performance, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing clumping in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (AUC of 50\%) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).

@@

First, in scenario \textnumero1, we also tested the T-Trees method. 
For example, as compared to the ``logit-simple'' method, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because it uses more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}). Even when simulating a fancy model in which there are dominant and interaction effects that decision trees based methods should handle, AUC is still lower with T-Trees as compared with the ``logit-simple'' method.

>>>>

We present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in a previous study. We quickly discard the T-Trees method and show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in previous studies.\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:}
}

@@

Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs have proven to be useful for different applications, such as finding a common genetic contribution between two diseases \cite[]{purcell2009common}. 
Personalized medicine is another major application of PRSs. Personalized medicine ..will.. use PRSs in screening campaigns in order to identify high-risk individuals for a given disease. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
Moreover, screening based on sequencing seems to make individuals positively change their health behavior, while neither causing patient anxiety nor depression \cite[]{vassy2017impact}.
Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. 
For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.

Many methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- also sometimes called genetic profiling ..or even just PRS.. -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. 
This technique only ..uses/resorts to.. summary statistics, which makes it very ..suitable.. and also very fast.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. 
Yet, these models are not designed for predicting a binary trait such as a disease status and have proven not to be optimal at such ..a.. task in another comparative study \cite[]{abraham2013performance}. 
Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRSs for complex human disease by jointly estimating SNP effects. Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.

@@

There have already been comparative studies on predicting human binary phenotypes based on genotype data. 
\citep{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
\cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models performed better in a range of sparse to polygenic disease architectures. 
\cite{spiliopoulou2015genomic} investigated the effect of relatedness in prediction performance.
\cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but ..richer/more extenstive.. than the one used in \cite{zhou2013polygenic}. In order to make our comparison as comprehensive as possible, we compare different disease architectures (number, size and location of causal effects and heritability) with different model for generating liability ..scores.., one with only linear effects, and one which combines linear, dominant and ..interaction/interaction-type.. effects. First, we quickly discard T-Trees as being the ..best.. method for predicting disease status based on genotypes.
Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in \cite{ware2017heterogeneity}.

@@

We use real genotypes of European individuals from a case/control celiac disease cohort \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}.
Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}.
As for simulations presented here, in order to remove the structure induced by the celiac disease status, we ..first keep only/select/restrict/retain.. controls from this cohort. 
Secondly, in order to ..alleviate.. population structure, we ..keep only.. people from the UK. Thirdly, we further subset these British people based on deviation from the robust Mahalanobis distance ..on/for.. principal components. This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing''). 
Note that we decided to ..alleviate.. population structure because it can affect the predictive performance of methods, which we will investigate in another study.

@@

In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% [CITATION?]. We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set ($\{1000, 2000, \dots, 6000\}$), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part/component/term.. of the phenotypes (a ``simple'' model with only linear effects, and a ``fancy'' model which combines linear, dominant and ..interaction/interaction-type.. effects).
For the ``simple'' model, we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights (generated with a Gaussian or a Laplace distribution), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs). 
For the ``fancy'' model, we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$). 
We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ 
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$. 
Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still ..display/have.. a marginal effect. 
For both models, the ..$y_i$/values of $y_i$.. are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.

We ..make/construct/design/implement/work with.. 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000. It compares all methods described in section \ref{sec:methods} ..on/for/along/in relation to.. all other combinations of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}). Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more than one hour of computation for a single simulation only. 
Scenario \textnumero2 consists of 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of chromosome 6 only. Reducing the number of SNPs aims at increasing the polygenicity of the simulated models and at ..virtually/artificially.. increasing the sample size [REF? DUDBRIDGE?]. For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of the PRS and the ``logit-simple'' methods. 100 simulations are run using 300 causal SNPs anywhere on the genome.

@@

In this study, we use two different measures of predictive accuracy. 
First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,hanley1982meaning,lusted1971signal}. In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
This measure indicates the extent to which we can distinguish between cases and controls using PRSs.
As a second measure, the partial AUC for specificities between 90\% and 100\% is also reported \cite[]{dodd2003partial,mcclish1989analyzing}. This measure is similar to the AUC, but focuses on high specificities (i.e.\ discarding most of the healthy population as being diseased), which is often the most useful part of the ROC curve in clinical settings.

@@

\subsection{Methods compared} \label{sec:methods}

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Studies (GWAS), referred to as summary statistics.
For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another. 
Then, thresholding (T) consists in removing SNPs that are within a certain level of significance (p-value threshold to be determined). 
Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T, E_\text{clumping}) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.

For the C+T method, we report three different scores of prediction: one including all the SNPs (remaining after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}). 
Note that, normally, this optimal threshold would have to be learned based on the training set only in order to prevent from over-fitting the data. We consider this reported maximum AUC as an upper bound of the AUC for the C+T method.
We call these three reported scores respectively ``PRS-all'', ``PRS-stringent'' and ``PRS-max'' in the results.

T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}. Since this method is ..significantly.. different from the other ones presented here and give that it achieved surprisingly high predictive performance in the past, we decided to include T-Trees in this comparison. We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees only (instead of 1000) because ..it.. showed only a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown). %We call this method ``T-Trees'' in the results.

The last method ..in/of.. this comparison is a penalized logistic regression.
We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined. 
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be exactly ..zero/equal to zero.. and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$); 
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we always use the elastic-net regularization with $\alpha = 0.5$, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}. 
This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values). To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). 
Secondly, in turn, each fold is considered as an inner validation set and the other ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is selected. Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ by using the geometric median) or resulting scores can be combined (e.g.\ by averaging the $K$ probabilities for each individual). 
We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.
Because of the L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We refer to this method as ``logit-simple'' in the results section.

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times as many variables as the initial one, on which we can apply the penalized logistic regression with the CMSA procedure, as described previously. We call this method ``logit-triple'' in the results.

@@

When reporting results, we use estimates of maximum achievable AUC values for the simulations: 84\% (94\%) for heritabilities of 50\% (80\%) based on three estimations (see Supplementary Materials). Note that for the simulations we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).

@@

In scenario \textnumero1, we compared our penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') on the whole genome (Table \ref{tab:simus}). 
Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the C+T method in any of the scenarios (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}). 
The penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}). 
In the case where there are correlations between predictors (i.e.\ when we simulate 30 SNPs in the HLA region), the ``logit-simple'' method performs particularly ..better.. than the ``PRS-max'' one, with a mean AUC of 92.5\% versus 83\%, for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
We also report that the choice of the threshold of inclusion of SNPs in the C+T method can impact its performance greatly, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing the clumping phase in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (AUC of 50\%) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).

@@

First, in scenario \textnumero1, we also tested the T-Trees method. 
For example, as compared to the ``logit-simple'' method, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}). Even when simulating a fancy model in which there are dominant and ..interaction/interaction-type.. effects that decision trees based methods should be able to handle, AUC is still lower with T-Trees as compared with the ``logit-simple'' method.

Secondly, as for the comparison between the two penalized logistic regressions in scenario \textnumero1, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}). Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``logit-simple''.
